An Open-Source Implementation of the CMPS Algorithm for Assessing Similarity of Bullets

### 2022-04-12 rj_format/rj-format-ju-hofmann/ju-hofmann.Rmd
change to rjtools

---
title: An Open-Source Implementation of the CMPS Algorithm for Assessing Similarity of Bullets
header-includes:
  - \usepackage{amsmath}
  - \usepackage{xcolor}
  - \usepackage{longtable}
  - \usepackage{adjustbox}            
  - \usepackage[font=small,skip=5pt]{caption}
  - \usepackage{subcaption}
  - \usepackage{afterpage}
  - \usepackage{bbm}
author:
  # see ?rjournal_article for more information
  - name: Wangqian Ju
    affiliation: Department of Statistics
    address:
    - Center for Statistics and Applications in Forensic Evidence
    - Iowa State University
    - 2438 Osborn Dr
    - Ames, IA 50011
    url: https://github.com/willju-wangqian
    email:  wju@iastate.edu
    orcid: 0000-0002-9977-377X
  - name: Heike Hofmann
    url: https://github.com/heike
    email: hofmann@iastate.edu
    orcid: 0000-0002-9079-593X
    affiliation: Department of Statistics
    address:
    - Center for Statistics and Applications in Forensic Evidence
    - Iowa State University
    - 2438 Osborn Dr
    - Ames, IA 50011
thanks: |
  This work was partially funded by the Center for Statistics and Applications in Forensic Evidence (CSAFE) through Cooperative Agreement 70NANB20H019 between NIST and Iowa State University, which includes activities carried out at Carnegie Mellon University, Duke University, University of California Irvine, University of Virginia, West Virginia University, University of Pennsylvania, Swarthmore College and University of Nebraska, Lincoln.    
abstract: >
  In this paper, we introduce the R package \pkg{cmpsR}, an open-source implementation of the Congruent Matching Profile Segments (CMPS) method developed at the National Institute of Standards and Technology (NIST) for objective comparison of striated tool marks. The functionality of the package is showcased by examples of bullet signatures that come with the package. Graphing tools are implemented in the package as well for users to assess and understand the CMPS results. Initial tests were performed on bullet signatures generated from two sets of 3D scans in the Hamby study under the framework suggested by the R package \pkg{bulletxtrctr}. New metrics based on CMPS scores are introduced and compared with existing metrics. A measure called Sum of squares ratio is included, and how it can be used for evaluating different scans, metrics, or parameters is showcased with the Hamby study data sets. An open-source implementation of the CMPS algorithm makes the algorithm more accessible, generates reproducible results, and facilitates further studies of the algorithm such as methods comparison.
preamble: |
  % Any extra LaTeX you need in the preamble

bibliography: ju-hofmann.bib
output: rticles::rjournal_article
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: sentence
---

```{=tex}
\newcommand{\hh}[1]{{\textcolor{orange}{#1}}}
\newcommand{\wj}[1]{{\textcolor{blue}{#1}}}
\setlength{\parindent}{0pt}
```

@@

```{r setup_knitr, echo=FALSE}
## From Josh O'Brien's stackoverflow answer:
## http://stackoverflow.com/questions/11030898/knitr-how-to-align-code-and-plot-side-by-side
## These two settings control text width in codefig vs. usual code blocks
partWidth <- 35
fullWidth <- 60
options(width = fullWidth)

@@

In this paper we present an open-source implementation of the algorithm of the Congruent Matching Profile Segments (CMPS) method.
@cmps developed the CMPS method for "objective comparison of striated tool marks" and demonstrated its use in some examples of comparing bullet signature correlations.
Although @cmps conceptually describe the CMPS algorithm in their paper, the authors did not release an implementation of their method.
Thus, our effort here is to introduce the CMPS method to the `R` [@R] community and provide an open-source, publicly available implementation of the algorithm to use, review, and improve.
Our implementation is made available as part of the `R` package \pkg{cmpsR} on CRAN.

According to the Uniform Crime Reporting Program of FBI (Federal Bureau of Investigation), "more than 76 percent (76.7) of the homicides for which the FBI received weapons data in 2020 involved the use of firearms" [@fbiucr]. At the same time, the number of murder victims jumped between 2019 and 2020 by more than 23 percent (23.4) to 17754 [@fbiucr20]. This increase is unprecedented  and highlights the important role that  firearm examination plays in all these cases. 
<!-- homicides in 2019: 14391; in 2020: 17754 -->
An important task of firearm examination is to answer the question of whether two pieces of evidence come from the same source or whether a piece of evidence matches a sample obtained from a specific firearm.
Here, in particular, we are interested to determine whether two bullets were fired from the same gun barrel.
Assessing the similarity between two bullets is based on a comparison of striation marks acquired during the firing process as bullets are propelled through the barrel.
Current state of art sees firearms examiners make an assessment of similarity based on a visual comparison, generally, using a comparison microscope [@afte].
This practice has been criticized for its lack of objectivity and the associated problem of determining valid error rates [@pcast].

A report published by the @nrc states that "[m]uch forensic evidence-including, for example, bite marks and firearm and toolmark identification---is introduced in criminal trials without any meaningful scientific validation, determination of error rates, or reliability testing to explain the limits of the discipline." To overcome those criticisms and concerns, researchers have been making an effort to build databases and develop frameworks and algorithms that bring an objective and quantitative assessment into the field.
The database of @nistdb with digital 3D topographic scans from various studies by @brundage, @hamby, @Hamby:2019 and others, provides a great resource for researchers. Algorithms that can be validated and tested and generate quantitative results are developed for 2D and 3D surface texture [@song2005], tool marks [@ChumbleyL_Scott2010VoTM], and striae on Land Engraved Areas (LEAs) of bullets [@aoas, @pmid30444940, @cmps]. 
<!--Note that this is not an exhaustive list of these efforts.-->
The notion of bullet signatures is one of the results of these efforts, and we will use it to explain the CMPS algorithm and showcase the \pkg{cmpsR} implementation.
Bullet signatures can be extracted from bullet LEAs, typically one for each land engraved area, and how those bullet signatures are extracted from scans will be discussed in the Background section.
Bullet signatures capture striation marks on the bullet in a numeric format and therefore serve as the foundation for algorithms. These scans played an important role in bringing objectivity into the field.
In the following sections we will: review the background of bullet signature comparisons, discuss how we followed the idea described by @cmps for the implementation of the CMPS algorithm, propose new metrics based on the CMPS score and a principled evaluation framework for algorithmic results comparison,  and present results of applying our implementation to real data.

@@

The datasets we worked with come from the James Hamby Consecutively Rifled Ruger Barrel Study [@brundage; @hamby; @Hamby:2019], in particular, Hamby set 252 and Hamby set 44.
For each Hamby set, a total of 35 bullets is fired from (the same) ten consecutively manufactured Ruger P-85 pistol barrels.
Two bullets are fired from each barrel, making up a set of 20 reference bullets.
An additional 15 bullets are fired from these ten barrels in a fashion unknown to the study participant.
The aim of the Hamby Study was to have firearms examiners identify which barrel each of the 15 questioned bullets was fired from.
The Ruger P-85 barrels are traditionally rifled barrels with six grooves and lands as shown in \autoref{fig:bullet}.
During the firing process, grooves and lands are engraved on a bullet.
Firearms examiners use striation marks on land engraved areas (LEAs) for their visual comparison.
For algorithmic purposes, 3D topographical images of land engraved areas were obtained and stored in x3p format (XML 3-D Surface Profile).
The x3p format provides a standard way of exchanging 2D and 3D profile data.
It conforms to [the ISO5436-2 standard](http://sourceforge.net/p/open-gps/mwiki/X3p/) adopted by the OpenFMC (Open Forensic Metrology Consortium), a group of firearm forensics researchers who contributes to the establishment of best practices of using metrology in forensic science.
Hamby set 252 was scanned using a NanoFocus lens at 20x magnification with the scan resolution being 1.5625 \textmu m $\times$ 1.5625 \textmu m per pixel.
Hamby set 44 was scanned at the Roy J Carver High-Resolution microscopy lab at Iowa State.
These scans were acquired with a Sensofar Confocal Light Microscope at 20x magnification for a nominal resolution of 0.645 \textmu m $\times$ 0.645 \textmu m per pixel.
Both Hamby set 252 and Hamby set 44 are publicly available from the NIST Ballistics Database Project [@nistdb].

@@

The automated framework for extracting signatures from x3p files used in this paper was proposed by @aoas and is implemented in the `R` packages \pkg{x3ptools} [@x3ptools] and \pkg{bulletxtrctr} [@bulletxtrctr].
\pkg{x3ptools} is a package to read, write, and generally, process x3p files.
The \pkg{bulletxtrctr} package implements a pipeline for extracting and comparing signatures from scans of land-engraved areas.

\autoref{fig:process} gives an overview of all of the steps in the process from scan to signatures.
\autoref{fig:process}(a) shows a rendering of a 3D scan of a bullet land engraved area.
The raised portion of the surface on the left and right of the scan are parts of the adjacent groove engraved areas (GEAs), the middle area shows a land-engraved area (LEA) with well expressed striation marks.
The first step of obtaining the bullet signature is to extract a cross-section at a fixed height along the land engraving.

The thin white horizontal line in \autoref{fig:process}(a) indicates which cross-section was identified by the algorithm to represent the LEA; \autoref{fig:process}(b) shows the corresponding cross-sectional view.
Groove engraved areas are removed from the analysis as indicated by the vertical blue lines in \autoref{fig:process}(c) and \autoref{fig:process}(d)).
A non-parametric LOESS smooth [@loess] is fitted to capture the bullet curvature (\autoref{fig:process}(e)) and, finally, the **bullet signature** (\autoref{fig:process}(f)) is obtained as residuals of the cross-section and the fitted smooth.
Note, that in @cmps bullet signatures are referred to as bullet profiles.
However, to avoid confusion, we distinguish the notion of bullet signatures from bullet profiles.
Bullet profiles are shown in panels (b), (c), and (d) of \autoref{fig:process}, while \autoref{fig:process}(f) shows the corresponding bullet signature.
Identifying the groove engraved areas correctly is fundamental for a correct down-stream analysis of the signatures.
To allow for a human-in-the middle inspection and intervention we have provided an interactive web-application, implemented as `R` Shiny App [@shiny], \pkg{bulletinspectR} to identify and correct those errors.
An example of the extraction process with corresponding code and parameter settings can be found in the "Supplementary materials". Note that the process of extracting signatures might be different from the one used in @cmps because no code or parameter settings are made available publicly.

@@

```{r process, echo=FALSE, out.width=".9\\textwidth", fig.cap="A framework of obtaining a bullet signature. (a) rendering from the 3D topographic scan of a land engraved area (LEA). The selected crosscut location is indicated by a thin white horizontal line. (b) view of the cross-section of the land engraved area at the white line in (a). (c) the crosscut data plotted in 2D; blue vertical lines indicate the position of left and right grooves. (d) the crosscut data after chopping the left and right grooves. (e) the fitted curvature using LOESS. (f) after removing the curvature from the crosscut data, the bullet signature is obtained"}
knitr::include_graphics("ju-hofmann_files/figure-latex/figure1_v2.png", dpi = 144)
```

@@

In this example, the comparison between \code{sig1} and \code{sig2}, two signatures coming from the same source (a known-match comparison), gets a CMPS score of 17; the comparison between \code{sig1} and \code{sig3}, two signatures coming from different sources (a known non-match comparison), gets a CMPS score of 1.

@@

The section "Implementation" will go through the algorithm and show how to use the \pkg{cmpsR} package. 
A further example that illustrates the main points is also included.
In the section on 'Evaluation metrics" we introduce new CMPS metrics that summarize land-level CMPS scores and a sum of squares ratio that can be used to evaluate algorithmic results.
The section "Results" presents the results of evaluating the \pkg{cmpsR} package using Hamby set 252 and Hamby set 44. The results from Hamby set 252 are used to verify that our implementation is, at least qualitatively, comparable to the algorithm described in @cmps. Results from Hamby 44 show the need for a further investigation of the parameter choices even in the case of bullets fired from the same barrels. 
The last section covers some final discussion and conclusions.

@@

1.  **cut the comparison signature into consecutive, non-overlapping and equal-length basis segments:** The command \code{get\_segs(x, len=50)} implements this step: it takes bullet signature \code{x} in the format of a numeric vector and cuts it into consecutive, non-overlapping and equal-length segments of length \code{len}, which are referred to as "basis segments". Note that the parameter \code{len} determines the length of a basis segment and thus affects the total number of basis segments, which is the upper limit of the CMPS score of a comparison. The default value of \code{len} is 50, which will result in about 25 basis segments for the example data of the package.

2.  **identify candidate positions:** For each basis segment a set of candidate registration positions on the comparison signature is identified based on the segment's similarity to the reference signature.
    In a first step, the cross-correlation function of the segment to the reference is calculated, then a number of positions with high correlation values are identified as candidate positions.
    In case multiple segment lengths are considered, the length of each basis segment is expanded (by default it is doubled) and these two steps are repeated.
    Only when candidate positions coincide (or are similar enough), they are considered further.
    \autoref{fig:segplots} and \autoref{fig:seg_all} illustrate these ideas.

    -   **Calculate the cross-correlation curve:** Calculate the cross-correlation curve between a basis segment \code{x} and the reference signature \code{y} using the function \code{get\_ccf4(x, y, ...)} as shown in \autoref{fig:segplot2}. The position indicates the lag by which a basis segment is moved with respect to its original placement. A position is considered "good" if it results in a peak in the cross-correlation between the basis segment and the reference.

    -   **Correlation peaks:** Two strategies referred to as "multi-peak inspection" and "multi-peak inspection at different segment lengths" in @cmps are used for identifying positions of correlation peaks as candidate positions. The latter is also called "multi-segment lengths strategy". The parameter \code{npeaks\_set} in \code{extract\_feature\_cmps(...)} determines which strategy to use and the number of candidate positions:

        -   If \code{npeaks\_set} is an integer vector of length 1, for example \code{npeaks\_set = 5}, the positions of the top five peaks in the cross-correlation curve are identified as candidate positions for registration.
        -   If \code{npeaks\_set} is an integer vector of length more than 1, for example \code{npeaks\_set = c(5, 3, 1)}, the multi-segment lengths strategy will be used: calculate the cross-correlation function between a basis segment and the reference and identify positions of the top five peaks; double the segment length to a specified value, re-calculate the cross-correlation function, and identify three peaks; repeat this process and identify a single peak in the newly computed cross-correlation function. \autoref{fig:seg_all} shows an example of three levels of basis segment 6 and their corresponding cross-correlation curves and identified peaks. Note that in @cmps the segment length is doubled at each level of a basis segment, but in the present implementation users are allowed to choose the segment length at each level.
        -   \code{get\_ccr\_peaks(comp, segments, seg\_outlength, nseg = 1, npeaks = 5)} computes the cross-correlation curve between a basis or increased segment and the reference signature and finds peaks in the cross-correlation curve. The number of peaks detected is equal to \code{npeaks}, which is an integer. \code{segments}, \code{seg\_outlength}, and \code{nseg} determines the segment in the cross-correlation computation, and \code{comp} gives the reference signature. If the multi-segment lengths strategy is used, then \code{get\_ccr\_peaks(...)} is called in a \code{lapply()} for each level of the basis segment. The resulting list is called \code{ccr\_list}.

    -   **multi-segment lengths strategy:** with the multi-segment lengths strategy being used, a position is identified as a candidate position for registration and is called as a "consistent correlation peak" if it results in a top peak in the cross-correlation curve with a tolerance zone determined by \code{Tx} in all segment levels. Note that in @cmps, a segment at its largest scale (highest level) always identifies one peak, but we do not have this requirement in our implementation.

        -   the function \code{get\_seg\_scale(segments, nseg, out\_length)} is used to obtain the (potentially increased) version of a basis segment. \code{segments}, which is a list containing all basis segments generated by the function \code{get\_segs(...)} in step 1, and \code{nseg} are used to determine the basis segment to be increased. \code{out\_length} specifies the length of the output segment.
        -   \code{get\_ccp(ccr\_list, Tx = 25)} tries to identify the "consistent correlation peak". \code{ccr\_list} is the result of \code{lapply()} and \code{get\_ccr\_peaks(...)}, and \code{Tx} determines the size of a tolerance zone used in identifying the consistent correlation peak. \code{get\_ccp(...)} returns \code{NULL} if there is no consistent correlation peak.

3.  **determine the congruent registration position:** A candidate position "receives" votes from basis segments that identify it or a close position within a tolerance zone of \code{Tx} as a candidate position in step 2.
    Votes for all candidate positions are tallied, and the position with the highest number of votes gets chosen as the *congruent registration position*, indicating that most of the basis segments find their highly similar counterpart in the reference signature in terms of correlation at this registration position.
    In case of ties the middle position is taken as the congruent registration position.
    Basis segments with a congruent registration position are called "congruent matching profile segments" (CMPS).
    The total number of CMPS is the CMPS score of the comparison.
    \code{get\_CMPS(input\_ccp, Tx = 25)} is the function that tallies the votes and determines the congruent registration position and congruent matching profile segments (CMPS).

Note that there are several parameters in the CMPS algorithm that will affect the final results and are left to the users to decide, such as the length of basis segments \code{seg\_length} in step 1, the number of peaks \code{npeaks\_set} identified on each level in step 2, and the length of the tolerance zone \code{Tx} in both step 2 and 3.
In our implementation of the CMPS algorithm we used the parameters given in the original CMPS paper [@cmps] as the default values for these parameters.
However, the authors state that no cross-validation has been done - there might also be issues with respect to resolution of the scans.
Further research is needed, until then users are advised to think of default values as starting values and consider alternatives. However, the evaluation framework based on the Sum of Squares Ratio introduced in the later section could be used to evaluate the choices of these parameters. 
The main function that combines all steps in the CMPS algorithm described above is called \code{extract\_feature\_cmps(...)}.
Here we present it with its default parameters.

@@

The function \code{extract\_feature\_cmps} allows for the following input from users besides the previously discussed parameters \code{seg\_length}, \code{npeaks\_set}, and \code{Tx}:

-   \code{x} and \code{y} are two signatures: \code{x} serves as the comparison signature (which will be divided into basis segments) and \code{y} is the reference signature;

>>>>

---
title: "An Open-Source Implementation of the CMPS Algorithm for Assessing Similarity of Bullets"
header-includes:
  - \usepackage{amsmath}
  - \usepackage{xcolor}
  - \usepackage{longtable}
  - \usepackage{adjustbox}            
  - \usepackage[font=small,skip=5pt]{caption}
  - \usepackage{subcaption}
  - \usepackage{afterpage}
  - \usepackage{bbm}
author:
  # see ?rjournal_article for more information
  - name: Wangqian Ju
    affiliation: Department of Statistics
    address:
    - Center for Statistics and Applications in Forensic Evidence
    - Iowa State University
    - 2438 Osborn Dr
    - Ames, IA 50011
    url: https://github.com/willju-wangqian
    email:  wju@iastate.edu
    orcid: 0000-0002-9977-377X
  - name: Heike Hofmann
    url: https://github.com/heike
    email: hofmann@iastate.edu
    orcid: 0000-0002-9079-593X
    affiliation: Department of Statistics
    address:
    - Center for Statistics and Applications in Forensic Evidence
    - Iowa State University
    - 2438 Osborn Dr
    - Ames, IA 50011
abstract: >
  In this paper, we introduce the R package \pkg{cmpsR}, an open-source implementation of the Congruent Matching Profile Segments (CMPS) method developed at the National Institute of Standards and Technology (NIST) for objective comparison of striated tool marks. The functionality of the package is showcased by examples of bullet signatures that come with the package. Graphing tools are implemented in the package as well for users to assess and understand the CMPS results. Initial tests were performed on bullet signatures generated from two sets of 3D scans in the Hamby study under the framework suggested by the R package \pkg{bulletxtrctr}. New metrics based on CMPS scores are introduced and compared with existing metrics. A measure called Sum of squares ratio is included, and how it can be used for evaluating different scans, metrics, or parameters is showcased with the Hamby study data sets. An open-source implementation of the CMPS algorithm makes the algorithm more accessible, generates reproducible results, and facilitates further studies of the algorithm such as methods comparison.
thanks: |
  This work was partially funded by the Center for Statistics and Applications in Forensic Evidence (CSAFE) through Cooperative Agreement 70NANB20H019 between NIST and Iowa State University, which includes activities carried out at Carnegie Mellon University, Duke University, University of California Irvine, University of Virginia, West Virginia University, University of Pennsylvania, Swarthmore College and University of Nebraska, Lincoln.    
preamble: |
  % Any extra LaTeX you need in the preamble
bibliography: ju-hofmann.bib
type: package
output: 
  rjtools::rjournal_web_article
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: sentence
---

@@

```{r setup-knitr, echo=FALSE}
## From Josh O'Brien's stackoverflow answer:
## http://stackoverflow.com/questions/11030898/knitr-how-to-align-code-and-plot-side-by-side
## These two settings control text width in codefig vs. usual code blocks
partWidth <- 35
fullWidth <- 60
options(width = fullWidth)

@@

In this paper, we present an open-source implementation of the algorithm of the Congruent Matching Profile Segments (CMPS) method.
@cmps developed the CMPS method for "objective comparison of striated tool marks" and demonstrated its use in some examples of comparing bullet signature correlations.
Although @cmps conceptually describe the CMPS algorithm in their paper, the authors did not release an implementation of their method.
Thus, our effort here is to introduce the CMPS method to the `R` [@R] community and provide an open-source, publicly available implementation of the algorithm to use, review, and improve.
Our implementation is made available as part of the `R` package \CRANpkg{cmpsR} on CRAN.

According to the Uniform Crime Reporting Program of the FBI (Federal Bureau of Investigation), "more than 76 percent (76.7) of the homicides for which the FBI received weapons data in 2020 involved the use of firearms" [@fbiucr]. At the same time, the number of murder victims jumped between 2019 and 2020 by more than 23 percent (23.4) to 17754 [@fbiucr20]. This increase is unprecedented  and highlights the important role that  firearm examination plays in all these cases. 
<!-- homicides in 2019: 14391; in 2020: 17754 -->
An important task of firearm examination is to answer the question of whether two pieces of evidence come from the same source or whether a piece of evidence matches a sample obtained from a specific firearm.
Here, in particular, we are interested to determine whether two bullets were fired from the same gun barrel.
Assessing the similarity between two bullets is based on a comparison of striation marks acquired during the firing process as bullets are propelled through the barrel.
The current state of art sees firearms examiners make an assessment of similarity based on a visual comparison, generally, using a comparison microscope [@afte].
This practice has been criticized for its lack of objectivity and the associated problem of determining valid error rates [@pcast].

A report published by the @nrc states that "[m]uch forensic evidence-including, for example, bite marks and firearm and toolmark identification---is introduced in criminal trials without any meaningful scientific validation, determination of error rates, or reliability testing to explain the limits of the discipline." To overcome those criticisms and concerns, researchers have been making an effort to build databases and develop frameworks and algorithms that bring an objective and quantitative assessment into the field.
The database of @nistdb with digital 3D topographic scans from various studies by @brundage, @hamby, @Hamby:2019 and others, provides a great resource for researchers. Algorithms that can be validated and tested and generate quantitative results are developed for 2D and 3D surface texture [@song2005], tool marks [@ChumbleyL_Scott2010VoTM], and striae on Land Engraved Areas (LEAs) of bullets [@aoas, @pmid30444940, @cmps]. 
<!--Note that this is not an exhaustive list of these efforts.-->
The notion of bullet signatures is one of the results of these efforts, and we will use it to explain the CMPS algorithm and showcase the \pkg{cmpsR} implementation.
Bullet signatures can be extracted from bullet LEAs, typically one for each land engraved area, and how those bullet signatures are extracted from scans will be discussed in the Background section.
Bullet signatures capture striation marks on the bullet in a numeric format and therefore serve as the foundation for algorithms. These scans played an important role in bringing objectivity into the field.
In the following sections, we will: review the background of bullet signature comparisons, discuss how we followed the idea described by @cmps for the implementation of the CMPS algorithm, propose new metrics based on the CMPS score and a principled evaluation framework for algorithmic results comparison,  and present results of applying our implementation to real data.

@@

The datasets we worked with come from the James Hamby Consecutively Rifled Ruger Barrel Study [@brundage; @hamby; @Hamby:2019], in particular, Hamby set 252 and Hamby set 44.
For each Hamby set, a total of 35 bullets is fired from (the same) ten consecutively manufactured Ruger P-85 pistol barrels.
Two bullets are fired from each barrel, making up a set of 20 reference bullets.
An additional 15 bullets are fired from these ten barrels in a fashion unknown to the study participant.
The aim of the Hamby Study was to have firearms examiners identify which barrel each of the 15 questioned bullets was fired from.
The Ruger P-85 barrels are traditionally rifled barrels with six grooves and lands as shown in Figure \@ref(fig:bullet).
During the firing process, grooves and lands are engraved on a bullet.
Firearms examiners use striation marks on land engraved areas (LEAs) for their visual comparison.
For algorithmic purposes, 3D topographical images of land engraved areas were obtained and stored in x3p format (XML 3-D Surface Profile).
The x3p format provides a standard way of exchanging 2D and 3D profile data.
It conforms to [the ISO5436-2 standard](http://sourceforge.net/p/open-gps/mwiki/X3p/) adopted by the OpenFMC (Open Forensic Metrology Consortium), a group of firearm forensics researchers who contributes to the establishment of best practices of using metrology in forensic science.
Hamby set 252 was scanned using a NanoFocus lens at 20x magnification with the scan resolution being 1.5625 \textmu m $\times$ 1.5625 \textmu m per pixel.
Hamby set 44 was scanned at the Roy J Carver High-Resolution microscopy lab at Iowa State.
These scans were acquired with a Sensofar Confocal Light Microscope at 20x magnification for a nominal resolution of 0.645 \textmu m $\times$ 0.645 \textmu m per pixel.
Both Hamby set 252 and Hamby set 44 are publicly available from the NIST Ballistics Database Project [@nistdb].

@@

The automated framework for extracting signatures from x3p files used in this paper was proposed by @aoas and is implemented in the `R` packages \CRANpkg{x3ptools} [@x3ptools] and \pkg{bulletxtrctr} [@bulletxtrctr].
\pkg{x3ptools} is a package to read, write, and generally, process x3p files.
The \pkg{bulletxtrctr} package implements a pipeline for extracting and comparing signatures from scans of land-engraved areas.

Figure \@ref(fig:process) gives an overview of all of the steps in the process from scan to signatures.
Figure \@ref(fig:process)(a) shows a rendering of a 3D scan of a bullet land engraved area.
The raised portion of the surface on the left and right of the scan are parts of the adjacent groove engraved areas (GEAs), the middle area shows a land-engraved area (LEA) with well expressed striation marks.
The first step of obtaining the bullet signature is to extract a cross-section at a fixed height on the land engraving.

The thin white horizontal line in Figure \@ref(fig:process)(a) indicates which cross-section was identified by the algorithm to represent the LEA; Figure \@ref(fig:process)(b) shows the corresponding cross-sectional view.
Groove engraved areas are removed from the analysis as indicated by the vertical blue lines in Figure \@ref(fig:process)(c) and Figure \@ref(fig:process)(d)).
A non-parametric LOESS smooth [@loess] is fitted to capture the bullet curvature ( Figure \@ref(fig:process)(e)) and, finally, the **bullet signature** (Figure \@ref(fig:process)(f)) is obtained as residuals of the cross-section and the fitted smooth.
Note, that in @cmps bullet signatures are referred to as bullet profiles.
However, to avoid confusion, we distinguish the notion of bullet signatures from bullet profiles.
Bullet profiles are shown in panels (b), (c), and (d) of Figure \@ref(fig:process), while Figure \@ref(fig:process)(f) shows the corresponding bullet signature.
Identifying the groove engraved areas correctly is fundamental for a correct downstream analysis of the signatures.
To allow for a human-in-the-middle inspection and intervention we have provided an interactive web application, implemented as `R` Shiny App [@shiny], \pkg{bulletinspectR} to identify and correct those errors.
An example of the extraction process with corresponding code and parameter settings can be found in the "Supplementary materials". Note that the process of extracting signatures might be different from the one used in @cmps because no code or parameter settings are made available publicly.

@@

```{r process, echo=FALSE, out.width=".9\\textwidth", fig.cap="A framework of obtaining a bullet signature. (a) rendering from the 3D topographic scan of a land engraved area (LEA). The selected crosscut location is indicated by a thin white horizontal line. (b) view of the cross-section of the land engraved area at the white line in (a). (c) the crosscut data plotted in 2D; blue vertical lines indicate the position of the left and right grooves. (d) the crosscut data after chopping the left and right grooves. (e) the fitted curvature using LOESS. (f) after removing the curvature from the crosscut data, the bullet signature is obtained."}
knitr::include_graphics("ju-hofmann_files/figure-latex/figure1_v2.png", dpi = 144)
```

@@

In this example, the comparison between `sig1` and `sig2`, two signatures coming from the same source (a known-match comparison), gets a CMPS score of 17; the comparison between `sig1` and `sig3`, two signatures coming from different sources (a known non-match comparison), gets a CMPS score of 1.

@@

The section "Implementation" will go through the algorithm and show how to use the \pkg{cmpsR} package. 
A further example that illustrates the main points is also included.
In the section on 'Evaluation metrics", we introduce new CMPS metrics that summarize land-level CMPS scores and a sum of squares ratio that can be used to evaluate algorithmic results.
The section "Results" presents the results of evaluating the \pkg{cmpsR} package using Hamby set 252 and Hamby set 44. The results from Hamby set 252 are used to verify that our implementation is, at least qualitatively, comparable to the algorithm described in @cmps. Results from Hamby 44 show the need for a further investigation of the parameter choices even in the case of bullets fired from the same barrels. 
The last section covers some final discussion and conclusions.

@@

1.  **cut the comparison signature into consecutive, non-overlapping, and equal-length basis segments:** The command `get_segs(x, len=50)` implements this step: it takes bullet signature `x` in the format of a numeric vector and cuts it into consecutive, non-overlapping and equal-length segments of length `len`, which are referred to as "basis segments". Note that the parameter `len` determines the length of a basis segment and thus affects the total number of basis segments, which is the upper limit of the CMPS score of a comparison. The default value of `len` is 50, which will result in about 25 basis segments for the example data of the package.

2.  **identify candidate positions:** For each basis segment a set of candidate registration positions on the comparison signature is identified based on the segment's similarity to the reference signature.
    In the first step, the cross-correlation function of the segment to the reference is calculated, then a number of positions with high correlation values are identified as candidate positions.
    In case multiple segment lengths are considered, the length of each basis segment is expanded (by default it is doubled) and these two steps are repeated.
    Only when candidate positions coincide (or are similar enough), they are considered further.
    Figure `r knitr::asis_output(ifelse(knitr::is_html_output(), '\\@ref(fig:segplots-web)', '\\@ref(fig:segplots)'))` and Figure `r knitr::asis_output(ifelse(knitr::is_html_output(), '\\@ref(fig:segplots-all-web)', '\\@ref(fig:segall)'))` illustrate these ideas.

    -   **Calculate the cross-correlation curve:** Calculate the cross-correlation curve between a basis segment `x` and the reference signature `y` using the function `get_ccf4(x, y, ...)` as shown in Figure `r knitr::asis_output(ifelse(knitr::is_html_output(), '\\@ref(fig:segplots-web)(b)', '\\@ref(fig:segplot2)'))`. The position indicates the lag by which a basis segment is moved with respect to its original placement. A position is considered "good" if it results in a peak in the cross-correlation between the basis segment and the reference.

    -   **Correlation peaks:** Two strategies referred to as "multi-peak inspection" and "multi-peak inspection at different segment lengths" in @cmps are used for identifying positions of correlation peaks as candidate positions. The latter is also called the "multi-segment lengths strategy". The parameter `npeaks_set` in `extract_feature_cmps(...)` determines which strategy to use and the number of candidate positions:

        -   If `npeaks_set` is an integer vector of length 1, for example, `npeaks_set = 5`, the positions of the top five peaks in the cross-correlation curve are identified as candidate positions for registration.
        -   If `npeaks_set` is an integer vector of length more than 1, for example, `npeaks_set = c(5, 3, 1)`, the multi-segment lengths strategy will be used: calculate the cross-correlation function between a basis segment and the reference and identify positions of the top five peaks; double the segment length to a specified value, re-calculate the cross-correlation function, and identify three peaks; repeat this process and identify a single peak in the newly computed cross-correlation function. Figure `r knitr::asis_output(ifelse(knitr::is_html_output(), '\\@ref(fig:segplots-all-web)', '\\@ref(fig:segall)'))` shows an example of three levels of basis segment 6 and their corresponding cross-correlation curves and identified peaks. Note that in @cmps the segment length is doubled at each level of a basis segment, but in the present implementation users are allowed to choose the segment length at each level.
        -   `get_ccr_peaks(comp, segments, seg_outlength, nseg = 1, npeaks = 5)` computes the cross-correlation curve between a basis or increased segment and the reference signature and finds peaks in the cross-correlation curve. The number of peaks detected is equal to `npeaks`, which is an integer. `segments`, `seg_outlength`, and `nseg` determine the segment in the cross-correlation computation, and `comp` gives the reference signature. If the multi-segment lengths strategy is used, then `get_ccr_peaks(...)` is called in a `lapply()` for each level of the basis segment. The resulting list is called `ccr_list`.

    -   **multi-segment lengths strategy:** with the multi-segment lengths strategy being used, a position is identified as a candidate position for registration and is called a "consistent correlation peak" if it results in a top peak in the cross-correlation curve with a tolerance zone determined by `Tx` in all segment levels. Note that in @cmps, a segment at its largest scale (highest level) always identifies one peak, but we do not have this requirement in our implementation.

        -   the function `get_seg_scale(segments, nseg, out_length)` is used to obtain the (potentially increased) version of a basis segment. `segments`, which is a list containing all basis segments generated by the function `get_segs(...)` in step 1, and `nseg` are used to determine the basis segment to be increased. `out_length` specifies the length of the output segment.
        -   `get_ccp(ccr_list, Tx = 25)` tries to identify the "consistent correlation peak". `ccr_list` is the result of `lapply()` and `get_ccr_peaks(...)`, and `Tx` determines the size of a tolerance zone used in identifying the consistent correlation peak. `get_ccp(...)` returns `NULL` if there is no consistent correlation peak.

3.  **determine the congruent registration position:** A candidate position "receives" votes from basis segments that identify it or a close position within a tolerance zone of `Tx` as a candidate position in step 2.
    Votes for all candidate positions are tallied, and the position with the highest number of votes gets chosen as the *congruent registration position*, indicating that most of the basis segments find their highly similar counterpart in the reference signature in terms of correlation at this registration position.
    In the case of ties, the middle position is taken as the congruent registration position.
    Basis segments with a congruent registration position are called "congruent matching profile segments" (CMPS).
    The total number of CMPS is the CMPS score of the comparison.
    `get_CMPS(input_ccp, Tx = 25)` is the function that tallies the votes and determines the congruent registration position and congruent matching profile segments (CMPS).

Note that there are several parameters in the CMPS algorithm that will affect the final results and are left to the users to decide, such as the length of basis segments `seg_length` in step 1, the number of peaks `npeaks_set` identified on each level in step 2, and the length of the tolerance zone `Tx` in both step 2 and 3.
In our implementation of the CMPS algorithm, we used the parameters given in the original CMPS paper [@cmps] as the default values for these parameters.
However, the authors state that no cross-validation has been done - there might also be issues with respect to the resolution of the scans.
Further research is needed, until then users are advised to think of default values as starting values and consider alternatives. However, the evaluation framework based on the Sum of Squares Ratio introduced in the later section could be used to evaluate the choices of these parameters. 
The main function that combines all steps in the CMPS algorithm described above is called `extract_feature_cmps(...)`.
Here we present it with its default parameters.

@@

The function `extract_feature_cmps` allows for the following input from users besides the previously discussed parameters `seg_length`, `npeaks_set`, and `Tx`:

-   `x` and `y` are two signatures: `x` serves as the comparison signature (which will be divided into basis segments) and `y` is the reference signature;
